---
title: "Phylo_Practicum_R"
output: html_document
date: "2026-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

04 Analyzing results of gene tree estimation
##packages needed

```{r cars}
library(ape)
library(phangorn)
library(phytools)
library(ggplot2)

tree_files <-list.files(pattern="\\.raxml.bestTree$") #List all .bestTree files. $ ensures the end of the name

gene_trees<- list() # list with all the trees
class(gene_trees)<- "multiPhylo" #make it a multiphylo object for ease of use with other 

i<-1
for(tree_file in tree_files){ ##go thru each file and read the tree
  gene_trees[[i]]<- read.tree(tree_file)
  i<-i+1
}
```


```{r}
#Find the appropriate root taxon
root_taxa <- c("H_vulgare_HVens23", "Er_bonaepartis_TB1", "S_vavilovii_Tr279", "Ta_caputMedusae_TB2")

# Extract the first matching species for each tree
gene_tree_outgroup<- rep(NA,length(gene_trees))
for(i in 1:length(gene_trees)){
  gene_tree <- gene_trees[[i]]
  found_taxa <- root_taxa[root_taxa %in% gene_tree$tip.label]
  
  # Return the first one if it exists
  if (length(found_taxa) > 0) {
    gene_tree_outgroup[i]<-found_taxa[1]
  }
}

#check if any gene trees don't have an outgroup taxon
any(is.na(gene_tree_outgroup)) ## should return FALSE

#re-reroot all our gene trees by the respective outgroup
for(i in 1:length(gene_trees)){
  gene_trees[[i]]<- root(gene_trees[[i]],
                         outgroup = gene_tree_outgroup[i],
                         resolve.root=TRUE)
  gene_trees[[i]]<-chronos(gene_trees[[i]]) ## make ultrametric for nicer densitree
}
```

From here we can plot trees
```{r}
plot(gene_trees[[2]]) #plots gene tree two
```

From there we can use code below for better visualization
```{r}
# see which genes have which taxa
all_labels <- unlist(lapply(gene_trees, function(x) x$tip.label)) ##count how often each individual appears
df <- as.data.frame(table(all_labels) / length(gene_trees)) #organize as a nice data frame for plotting

ggplot(df, aes(x = all_labels, y = Freq)) +
  geom_col() +
  labs(x = "Individual", y = "Proportion")+
  theme(axis.text.x = element_text(angle = 90))
```

Let's plot it
```{r}
st<-superTree(gene_trees)
st<-root(st,"H_vulgare_HVens23",resolve.root = T)
plot(st)
```

05 Full concatenation visualization

```{r}
library(ape)
library(phangorn)
library(phytools)

tre = read.tree(file="triticeae_allindividuals_OneCopyGenes.fasta.raxml.bestTree")
plot(tre)
nodelabels()
```

Rooting tree at the outgroup clade
```{r}
rtre = root(tre,node = 51, resolve.root=TRUE)
plot(ladderize(rtre))
```

Compare to the authors
```{r}
tre2 = read.tree(file="../../../data/Wheat_Relative_History_Data_Glemin_et_al/MLtree_OneCopyGenes.tree")
plot(tre2)
rtre2 = root(tre2,node = 51, resolve.root=TRUE)
plot(ladderize(rtre2))

RF.dist(tre,tre2) # calculates RF distance between the two trees whcih should be 0 in this case
```
What is RF distance?

06 10Mb concatenation window visualization

Reproduction of figure 1B from the paper
```{r}
library(ape)
library(phangorn)
library(phytools)
library(ggplot2)

getwd() #Check the working directory. we want to be in the results/RAxML/10Mb-concatenation folder
#setwd("PathTo/results/RAxML/10Mb-concatenation") #replace PathTo with the correct path and run if not in the correct folder

tree_files <-list.files(pattern="\\.raxml.bestTree$") #List all .bestTree files. $ ensures the end of the name

trees<- list() # list with all the trees
class(trees)<- "multiPhylo" #make it a multiphylo object for ease of use with other 

i<-1
for(tree_file in tree_files){ ##go thru each file and read the tree
  trees[[i]]<- read.tree(tree_file)
  i<-i+1
}
```

Root all trees in "H_vulgare_HVens23
```{r}
#re-reroot all our gene trees by the respective outgroup
for(i in 1:length(trees)){
  trees[[i]]<- root(trees[[i]],
                         outgroup = "H_vulgare_HVens23",
                         resolve.root=TRUE)
  trees[[i]]<-chronos(trees[[i]]) ## make ultrametric for nicer densitree
}
```

Create consensus parsimony supertree
```{r}
st<-superTree(trees)
st<-root(st,"H_vulgare_HVens23",resolve.root = T)
plot(st)
```

Plot density tree
```{r}
densiTree(trees,consensus=st,scaleX=T,type='cladogram', alpha=0.1)
```

Comparing our density tree to theirs (it isn't quite the same and I don't have the modified copy)
```{r}
trees2 <- read.nexus("../../../data/Wheat_Relative_History_Data_Glemin_et_al/Densitree_OneCopyGenes.nex")


#re-reroot all our gene trees by the respective outgroup
for(i in 1:length(trees2)){
  trees2[[i]]<- root(trees2[[i]],
                         outgroup = "H_vulgare_HVens23",
                         resolve.root=TRUE)
  trees2[[i]]<-chronos(trees2[[i]]) ## make ultrametric for nicer densitree
}

st2<-superTree(trees2)
st2<-root(st2,"H_vulgare_HVens23",resolve.root = T)
plot(st2)

densiTree(trees2,consensus=st2,scaleX=T,type='cladogram', alpha=0.1)
```

07 Visualizing species generated with Supertriplets
```{r}
setwd("~/Desktop/phylo-practicum/glemin-wheat/results")

library(ape)
tre = read.tree(file="07-supertree.tre")
plot(tre)

rtre = root(tre,outgroup="H_vulgare_HVens23", resolve.root=TRUE)
plot(rtre)
```

Can infer a species level phylogeny with multispecies coalescence since their are multiple individuals per species. To do this, we first create a mapping from individual ID to species ID
```{r}
setwd("~/Desktop/phylo-practicum/glemin-wheat/code")
##First we get all the individual names
genes_dir <- "../data/Wheat_Relative_History_Data_Glemin_et_al/OneCopyGenes/"
gene_files<-paste(genes_dir,list.files(genes_dir,pattern='\\.aln$'),sep='')

all_individuals <- character()
i<-1
for(f in gene_files){
  headers <- rownames(read.dna(f, format = "fasta"))
  all_individuals <- unique(c(all_individuals, headers))
}
all_individuals <- sort(all_individuals) # Sort alphabetically for consistency
all_individuals
```

Creating a text file where the name of each individual just removes the tag at the eng of the name and we will have species all having the same name
```{r}
cleaned_individuals <- sub("_[^_]+$", "", all_individuals)

#map all individuals to species
mapping <- paste(all_individuals,cleaned_individuals)
writeLines(mapping, "../results/07-species_mapping.txt") ## write to file
```

Generating 07-species-tree-astral4.tre
```{r}
astral4 -i 04-all_gene_trees.tre -a 07-species_mapping.txt -o 07-species-tree-astral4.tre

```



Visualizing 07-species-tree-astral4.tre
```{r}
setwd("~/Desktop/phylo-practicum/glemin-wheat/results")
library(ape)
tre = read.tree(file="07-species-tree-astral4.tre")
plot(tre)
rtre = root(tre,outgroup="H_vulgare", resolve.root=TRUE)
plot(rtre)

```


08 Species tree visualizations

Here we are going to plot the full cat tree and supertree side by side

```{r}
setwd("~/Desktop/phylo-practicum/glemin-wheat/results")

library(ape)

tree1 = read.tree(file="triticeae_allindividuals_OneCopyGenes.fasta.raxml.bestTree")
tree2 = read.tree(file="07-supertree.tre")

tree1 = root(tree1,outgroup="H_vulgare_HVens23", resolve.root=TRUE)
tree2 = root(tree2,outgroup="H_vulgare_HVens23", resolve.root=TRUE)

# Suppose your trees are called tree1 and tree2
# First, ladderize both trees
tree1 <- ladderize(tree1)
tree2 <- ladderize(tree2[[2]])

# Make sure they have the same tip labels
common_tips <- intersect(tree1$tip.label, tree2$tip.label)

length(common_tips)
length(tree1$tip.label)
length(tree2$tip.label)

# Reorder the second tree to match the tip order of the first
tree2 <- reorder.phylo(tree2, order = "cladewise")
tree2 <- rotateConstr(tree2, tree1$tip.label) # rotate to match tree1 tip order

# Plot side by side
par(mfrow = c(1, 2))
plot(tree1, main = "Full concatenation", cex = 0.8)
plot(tree2, main = "Supertree", cex = 0.8)
```

```{r}
library(phangorn)
RF.dist(tree1, tree2) ## not zero!

BiocManager::install("ggtree")
```


```{r}
library(ape)
library(ggtree)
library(tidytree)
library(dplyr)
library(ggplot2)

species_colors <- c(
  "Ae_umbellulata" = "yellow",
  "Ae_caudata" = "orange",
  "Ae_comosa" = "darkorange3",
  "Ae_uniaristata" = "sienna4",
  "Ae_bicornis" = "purple",
  "Ae_longissima" = "pink",
  "Ae_sharonensis" = "mediumpurple1",
  "Ae_searsii" = "maroon2",
  "Ae_tauschii" = "red",
  "T_boeoticum" = "darkgreen",
  "T_urartu" = "green",
  "Ae_speltoides" = "blue4",
  "Ae_mutica" = "steelblue1"
)

tip_species <- sapply(tree1$tip.label, function(x) {
  parts <- strsplit(x, "_")[[1]]
  paste(parts[1:2], collapse = "_")
})

p <- ggtree(tree1)
df <- p$data
df$branch_color <- "black"

for (sp in names(species_colors)) {

  tips <- tree1$tip.label[tip_species == sp]
  node <- getMRCA(tree1, tips)

  if (!is.null(node)) {
    descendants <- offspring(tree1, node)   # <- FIXED
    target_nodes <- c(node, descendants)

    df$branch_color[df$node %in% target_nodes] <- species_colors[sp]
  }
}

ggtree(tree1) %<+% df +
  aes(color = branch_color) +
  geom_tree(size = 1) +
  scale_color_identity() +
  geom_tiplab()

# Loop over species to color clades
for(sp in names(species_colors)) {
  # Get tips belonging to this species
  tips <- tree1$tip.label[tip_species == sp]
  # Get MRCA node
  node <- getMRCA(tree1, tips)
  if(!is.null(node)) {
    p <- p + geom_hilight(node = node, fill = species_colors[sp], alpha = 0.3)
  }
}

# Add tip labels
p <- p + geom_tiplab()
p

p + geom_text2(aes(subset = !isTip, label = node), hjust = -0.3)

p + geom_text2(aes(subset = !isTip, label = node), hjust = -0.3)

```

Here's how to rotate nodes so that the tree matches that of the publications

```{r}
p <- ggtree(tree1)

p2 <- rotate(p, node = 55)
p2 <- rotate(p2, node = 62)
p2 <- rotate(p2, node = 50)
p2 <- rotate(p2, node = 49)
p2 <- rotate(p2, node = 69)
p2 <- rotate(p2, node = 83)
p2 <- rotate(p2, node = 88)
p2 <- rotate(p2, node = 71)
p2 <- rotate(p2, node = 74)


for(sp in names(species_colors)) {
  tips <- tree1$tip.label[tip_species == sp]
  node <- getMRCA(tree1, tips)  # this works on phylo object
  if(!is.null(node)) {
    p2 <- p2 + geom_hilight(node = node, fill = species_colors[sp], alpha = 0.3)
  }
}

p2 <- p2 + geom_tiplab()
p2

```

The next step is the DensiTree (figure 1B)
- Used each 10Mb windows to show genetree discordance

```{r}
