---
title: "Phylo_Practicum_R"
output: html_document
date: "2026-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

04 Analyzing results of gene tree estimation
##packages needed

```{r cars}
library(ape)
library(phangorn)
library(phytools)
library(ggplot2)

tree_files <-list.files(pattern="\\.raxml.bestTree$") #List all .bestTree files. $ ensures the end of the name

gene_trees<- list() # list with all the trees
class(gene_trees)<- "multiPhylo" #make it a multiphylo object for ease of use with other 

i<-1
for(tree_file in tree_files){ ##go thru each file and read the tree
  gene_trees[[i]]<- read.tree(tree_file)
  i<-i+1
}
```


```{r}
#Find the appropriate root taxon
root_taxa <- c("H_vulgare_HVens23", "Er_bonaepartis_TB1", "S_vavilovii_Tr279", "Ta_caputMedusae_TB2")

# Extract the first matching species for each tree
gene_tree_outgroup<- rep(NA,length(gene_trees))
for(i in 1:length(gene_trees)){
  gene_tree <- gene_trees[[i]]
  found_taxa <- root_taxa[root_taxa %in% gene_tree$tip.label]
  
  # Return the first one if it exists
  if (length(found_taxa) > 0) {
    gene_tree_outgroup[i]<-found_taxa[1]
  }
}

#check if any gene trees don't have an outgroup taxon
any(is.na(gene_tree_outgroup)) ## should return FALSE

#re-reroot all our gene trees by the respective outgroup
for(i in 1:length(gene_trees)){
  gene_trees[[i]]<- root(gene_trees[[i]],
                         outgroup = gene_tree_outgroup[i],
                         resolve.root=TRUE)
  gene_trees[[i]]<-chronos(gene_trees[[i]]) ## make ultrametric for nicer densitree
}
```

From here we can plot trees
```{r}
plot(gene_trees[[2]]) #plots gene tree two
```

From there we can use code below for better visualization
```{r}
# see which genes have which taxa
all_labels <- unlist(lapply(gene_trees, function(x) x$tip.label)) ##count how often each individual appears
df <- as.data.frame(table(all_labels) / length(gene_trees)) #organize as a nice data frame for plotting

ggplot(df, aes(x = all_labels, y = Freq)) +
  geom_col() +
  labs(x = "Individual", y = "Proportion")+
  theme(axis.text.x = element_text(angle = 90))
```

Let's plot it
```{r}
st<-superTree(gene_trees)
st<-root(st,"H_vulgare_HVens23",resolve.root = T)
plot(st)
```

05 Full concatenation visualization

```{r}
library(ape)
library(phangorn)
library(phytools)

tre = read.tree(file="triticeae_allindividuals_OneCopyGenes.fasta.raxml.bestTree")
plot(tre)
nodelabels()
```

Rooting tree at the outgroup clade
```{r}
rtre = root(tre,node = 51, resolve.root=TRUE)
plot(ladderize(rtre))
```

Compare to the authors
```{r}
tre2 = read.tree(file="../../../data/Wheat_Relative_History_Data_Glemin_et_al/MLtree_OneCopyGenes.tree")
plot(tre2)
rtre2 = root(tre2,node = 51, resolve.root=TRUE)
plot(ladderize(rtre2))

RF.dist(tre,tre2) # calculates RF distance between the two trees whcih should be 0 in this case
```
What is RF distance?

06 10Mb concatenation window visualization

Reproduction of figure 1B from the paper
```{r}
library(ape)
library(phangorn)
library(phytools)
library(ggplot2)

getwd() #Check the working directory. we want to be in the results/RAxML/10Mb-concatenation folder
#setwd("PathTo/results/RAxML/10Mb-concatenation") #replace PathTo with the correct path and run if not in the correct folder

tree_files <-list.files(pattern="\\.raxml.bestTree$") #List all .bestTree files. $ ensures the end of the name

trees<- list() # list with all the trees
class(trees)<- "multiPhylo" #make it a multiphylo object for ease of use with other 

i<-1
for(tree_file in tree_files){ ##go thru each file and read the tree
  trees[[i]]<- read.tree(tree_file)
  i<-i+1
}
```

Root all trees in "H_vulgare_HVens23
```{r}
#re-reroot all our gene trees by the respective outgroup
for(i in 1:length(trees)){
  trees[[i]]<- root(trees[[i]],
                         outgroup = "H_vulgare_HVens23",
                         resolve.root=TRUE)
  trees[[i]]<-chronos(trees[[i]]) ## make ultrametric for nicer densitree
}
```

Create consensus parsimony supertree
```{r}
st<-superTree(trees)
st<-root(st,"H_vulgare_HVens23",resolve.root = T)
plot(st)
```

Plot density tree
```{r}
densiTree(trees,consensus=st,scaleX=T,type='cladogram', alpha=0.1)
```

Comparing our density tree to theirs (it isn't quite the same and I don't have the modified copy)
```{r}
trees2 <- read.nexus("../../../data/Wheat_Relative_History_Data_Glemin_et_al/Densitree_OneCopyGenes.nex")


#re-reroot all our gene trees by the respective outgroup
for(i in 1:length(trees2)){
  trees2[[i]]<- root(trees2[[i]],
                         outgroup = "H_vulgare_HVens23",
                         resolve.root=TRUE)
  trees2[[i]]<-chronos(trees2[[i]]) ## make ultrametric for nicer densitree
}

st2<-superTree(trees2)
st2<-root(st2,"H_vulgare_HVens23",resolve.root = T)
plot(st2)

densiTree(trees2,consensus=st2,scaleX=T,type='cladogram', alpha=0.1)
```

07 Visualizing species generated with Supertriplets
```{r}
setwd("~/Desktop/phylo-practicum/glemin-wheat/results")

library(ape)
tre = read.tree(file="07-supertree.tre")
plot(tre)

rtre = root(tre,outgroup="H_vulgare_HVens23", resolve.root=TRUE)
plot(rtre)
```

Can infer a species level phylogeny with multispecies coalescence since their are multiple individuals per species. To do this, we first create a mapping from individual ID to species ID
```{r}
setwd("~/Desktop/phylo-practicum/glemin-wheat/code")
##First we get all the individual names
genes_dir <- "../data/Wheat_Relative_History_Data_Glemin_et_al/OneCopyGenes/"
gene_files<-paste(genes_dir,list.files(genes_dir,pattern='\\.aln$'),sep='')

all_individuals <- character()
i<-1
for(f in gene_files){
  headers <- rownames(read.dna(f, format = "fasta"))
  all_individuals <- unique(c(all_individuals, headers))
}
all_individuals <- sort(all_individuals) # Sort alphabetically for consistency
all_individuals
```

Creating a text file where the name of each individual just removes the tag at the eng of the name and we will have species all having the same name
```{r}
cleaned_individuals <- sub("_[^_]+$", "", all_individuals)

#map all individuals to species
mapping <- paste(all_individuals,cleaned_individuals)
writeLines(mapping, "../results/07-species_mapping.txt") ## write to file
```

Generating 07-species-tree-astral4.tre
```{r}
astral4 -i 04-all_gene_trees.tre -a 07-species_mapping.txt -o 07-species-tree-astral4.tre

```



Visualizing 07-species-tree-astral4.tre
```{r}
setwd("~/Desktop/phylo-practicum/glemin-wheat/results")
library(ape)
tre = read.tree(file="07-species-tree-astral4.tre")
plot(tre)
rtre = root(tre,outgroup="H_vulgare", resolve.root=TRUE)
plot(rtre)

```


08 Species tree visualizations

Here we are going to plot the full cat tree and supertree side by side

```{r}
setwd("~/Desktop/phylo-practicum/glemin-wheat/results")

library(ape)

tree1 = read.tree(file="triticeae_allindividuals_OneCopyGenes.fasta.raxml.bestTree")
tree2 = read.tree(file="07-supertree.tre")

tree1 = root(tree1,outgroup="H_vulgare_HVens23", resolve.root=TRUE)
tree2 = root(tree2,outgroup="H_vulgare_HVens23", resolve.root=TRUE)

# Suppose your trees are called tree1 and tree2
# First, ladderize both trees
tree1 <- ladderize(tree1)
tree2 <- ladderize(tree2[[2]])

# Make sure they have the same tip labels
common_tips <- intersect(tree1$tip.label, tree2$tip.label)

length(common_tips)
length(tree1$tip.label)
length(tree2$tip.label)

# Reorder the second tree to match the tip order of the first
tree2 <- reorder.phylo(tree2, order = "cladewise")
tree2 <- rotateConstr(tree2, tree1$tip.label) # rotate to match tree1 tip order

# Plot side by side
par(mfrow = c(1, 2))
plot(tree1, main = "Full concatenation", cex = 0.8)
plot(tree2, main = "Supertree", cex = 0.8)
```

```{r}
library(phangorn)
RF.dist(tree1, tree2) ## not zero!

BiocManager::install("ggtree")
```


```{r}
library(ape)
library(ggtree)
library(tidytree)
library(dplyr)
library(ggplot2)

species_colors <- c(
  "Ae_umbellulata" = "yellow",
  "Ae_caudata" = "orange",
  "Ae_comosa" = "darkorange3",
  "Ae_uniaristata" = "sienna4",
  "Ae_bicornis" = "purple",
  "Ae_longissima" = "pink",
  "Ae_sharonensis" = "mediumpurple1",
  "Ae_searsii" = "maroon2",
  "Ae_tauschii" = "red",
  "T_boeoticum" = "darkgreen",
  "T_urartu" = "green",
  "Ae_speltoides" = "blue4",
  "Ae_mutica" = "steelblue1"
)

tip_species <- sapply(tree1$tip.label, function(x) {
  parts <- strsplit(x, "_")[[1]]
  paste(parts[1:2], collapse = "_")
})

p <- ggtree(tree1)
df <- p$data
df$branch_color <- "black"

for (sp in names(species_colors)) {

  tips <- tree1$tip.label[tip_species == sp]
  node <- getMRCA(tree1, tips)

  if (!is.null(node)) {
    descendants <- offspring(tree1, node)   # <- FIXED
    target_nodes <- c(node, descendants)

    df$branch_color[df$node %in% target_nodes] <- species_colors[sp]
  }
}

ggtree(tree1) %<+% df +
  aes(color = branch_color) +
  geom_tree(size = 1) +
  scale_color_identity() +
  geom_tiplab()

# Loop over species to color clades
for(sp in names(species_colors)) {
  # Get tips belonging to this species
  tips <- tree1$tip.label[tip_species == sp]
  # Get MRCA node
  node <- getMRCA(tree1, tips)
  if(!is.null(node)) {
    p <- p + geom_hilight(node = node, fill = species_colors[sp], alpha = 0.3)
  }
}

# Add tip labels
p <- p + geom_tiplab()
p

p + geom_text2(aes(subset = !isTip, label = node), hjust = -0.3)

p + geom_text2(aes(subset = !isTip, label = node), hjust = -0.3)

```

Here's how to rotate nodes so that the tree matches that of the publications

```{r}
p <- ggtree(tree1)

p2 <- rotate(p, node = 55)
p2 <- rotate(p2, node = 62)
p2 <- rotate(p2, node = 50)
p2 <- rotate(p2, node = 49)
p2 <- rotate(p2, node = 69)
p2 <- rotate(p2, node = 83)
p2 <- rotate(p2, node = 88)
p2 <- rotate(p2, node = 71)
p2 <- rotate(p2, node = 74)


for(sp in names(species_colors)) {
  tips <- tree1$tip.label[tip_species == sp]
  node <- getMRCA(tree1, tips)  # this works on phylo object
  if(!is.null(node)) {
    p2 <- p2 + geom_hilight(node = node, fill = species_colors[sp], alpha = 0.3)
  }
}

p2 <- p2 + geom_tiplab()
p2

```

The next step is the DensiTree (figure 1B)
- Used each 10Mb windows to show genetree discordance


11 Hybrid detection with MSCQuartets and Visualization

```{r}
setwd("~/Desktop/plantpath875/glemin-wheat/results")
library(MSCquartets)

gtrees <- read.tree(file="09-all_gene_trees_snaq.tre")
tnames <- unique(unlist(lapply(gtrees, function(x) x$tip.label)))

QT=quartetTable(gtrees,tnames)
RQT=quartetTableResolved(QT)

pTable3=quartetTreeTestInd(RQT,"T3")
quartetTablePrint(pTable3[1:6,])

write.csv(pTable3, "11-mscquartets-ptable.csv")

```

Reproducing Figure 3A and 3B

```{r}
i = 1
dt = read.table(paste0("Size_10_Mb_Chrom_3_Pos_",i,"-out.txt"), header=TRUE)

# in command line: head(dt)
```

For some reason, it isn't possible to reproduce 3A and 3B as it has something to do with the output from above only having counts in the AAAA column. So, we will use MSCQuartets instead to try to reproduce a similar figure

```{r}
df = read.csv("11-mscquartets-ptable.csv", header=TRUE)
```

```{r}
library(dplyr)

# find the columns for each species
p1_cols   <- grep("^Ae_mutica", names(df), value = TRUE)
p2_cols     <- grep("^T_boeoticum", names(df), value = TRUE)
bicornis_cols <- grep("^Ae_bicornis", names(df), value = TRUE)


# filter rows that have exactly one "1" in each group
df_filtered <- df %>%
  filter(
    rowSums(select(., all_of(p1_cols))   == 1, na.rm = TRUE) == 1,
    rowSums(select(., all_of(p2_cols))     == 1, na.rm = TRUE) == 1,
    rowSums(select(., all_of(bicornis_cols)) == 1, na.rm = TRUE) == 1
  )

sum(df_filtered$p_T3<0.05)/length(df_filtered$p_T3)

```

```{r}
outdf = data.frame(p1 = "Ae_mutica", p2 = "T_boeoticum", hybrid = "Ae_bicornis", pval = df_filtered$p_T3)

```

```{r}
hybrids = c("Ae_longissima", "Ae_sharonensis", "Ae_searsii", "Ae_tauschii", "Ae_caudata", "Ae_umbellulata", "Ae_comosa", "Ae_uniaristata")

for(h in hybrids){
    hyb_cols <- grep(h, names(df), value = TRUE)
    tmp <- df %>%
    filter(
    rowSums(select(., all_of(p1_cols))   == 1, na.rm = TRUE) == 1,
    rowSums(select(., all_of(p2_cols))     == 1, na.rm = TRUE) == 1,
    rowSums(select(., all_of(hyb_cols)) == 1, na.rm = TRUE) == 1
    )
    tmpdf = data.frame(p1 = "Ae_mutica", p2 = "T_boeoticum", hybrid=h, pval=tmp$p_T3)
    outdf = rbind(outdf,tmpdf)
}

#Up until now, we have the code to reproduce the top right panel of Figure 3A, but if we want to do the whole Figure, we need to change the parents too.

parents1 = c("Ae_mutica", "Ae_speltoides")
parents2 = c("T_boeoticum", "T_urartu")
hybrids = c("Ae_bicornis", "Ae_longissima", "Ae_sharonensis", "Ae_searsii", "Ae_tauschii", "Ae_caudata", "Ae_umbellulata", "Ae_comosa", "Ae_uniaristata")

for(pp1 in parents1){
    for(pp2 in parents2){
        if(pp1 == "Ae_mutica" && pp2 == "T_boeoticum") next

        p1_cols   <- grep(pp1, names(df), value = TRUE)
        p2_cols     <- grep(pp2, names(df), value = TRUE)
        for(h in hybrids){
            hyb_cols <- grep(h, names(df), value = TRUE)
            tmp <- df %>%
                filter(
                rowSums(select(., all_of(p1_cols))   == 1, na.rm = TRUE) == 1,
                rowSums(select(., all_of(p2_cols))     == 1, na.rm = TRUE) == 1,
                rowSums(select(., all_of(hyb_cols)) == 1, na.rm = TRUE) == 1
                )
            tmpdf = data.frame(p1 = pp1, p2 = pp2, hybrid=h, pval=tmp$p_T3)
            outdf = rbind(outdf,tmpdf)
        }
    }
}


```

Making the violin plot

```{r}
library(ggplot2)

## Setting the hybrid order (x axis):
hybrid_order <- c(
  "Ae_bicornis", "Ae_longissima", "Ae_sharonensis",
  "Ae_searsii", "Ae_tauschii", "Ae_caudata",
  "Ae_umbellulata", "Ae_comosa", "Ae_uniaristata"
)

outdf$hybrid <- factor(outdf$hybrid, levels = hybrid_order)

hybrid_colors <- c(
  "Ae_bicornis"     = "blue3",
  "Ae_longissima"  = "hotpink",
  "Ae_sharonensis" = "darkorchid1",
  "Ae_searsii"     = "deeppink",
  "Ae_tauschii"    = "red",
  "Ae_caudata"     = "goldenrod1",
  "Ae_umbellulata" = "yellow",
  "Ae_comosa"      = "orange2",
  "Ae_uniaristata" = "sienna4"
)



ggplot(outdf, aes(x = hybrid, y = pval, fill = hybrid)) +
  geom_violin(
    trim = FALSE,
    scale = "width",
    color = NA
  ) +
  geom_boxplot(
    width = 0.12,
    outlier.shape = NA,
    alpha = 0.85
  ) +
  facet_grid(
    rows = vars(p2),
    cols = vars(p1)
  ) +
  scale_fill_manual(values = hybrid_colors, drop = FALSE) +
  scale_x_discrete(drop = FALSE) +
  geom_hline(
    yintercept = 0.05,
    linetype = "dotted",
    color = "black"
  ) +
  labs(
    x = "D focal species",
    y = "MSCQuartets p-value"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.major.x = element_blank(),
    legend.position = "none"
  )

```

